---
title: "ADNI Clinical Progression"
author: "ATRI Biostatistics"
header-includes:
  - \usepackage{longtable}
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ADNI Clinical Progression}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

# ADNI progression summaries using the ADNIMERGE R package

## Description

This package vignette demonstrates how to use the [ADNIMERGE](index.html) package to generate progression summaries and plots with [R](http://www.r-project.org/) and [knitr](http://yihui.name/knitr/).

Report generated <!--rinline format(Sys.Date(), "%A, %B %d, %Y") --> from [Progresssion.Rmd](https://bitbucket.org/adni/adni.bitbucket.io/src)

This report contains estimates of the ADNI conversion rates using a nonparametric multistate modeling approach (Nicole Ferguson, Somnath Datta, Guy Brock (2012). msSurv: An R Package for Nonparametric Estimation of Multistate Models. Journal of Statistical Software, 50(14), 1-24. [http://www.jstatsoft.org/v50/i14/](http://www.jstatsoft.org/v50/i14/)).


This report does not consider transition to EMCI at any followup visits. The label LMCI, when applied at followup visits, actually refer to any MCI (EMCI or LMCI).
      
Only subjects with at least one followup Diagnostic Summary are included in this report.

```{r  knitropts, results = 'hide', echo = FALSE, message = FALSE}
knitr::opts_chunk$set(fig.path='figure/progression-', echo = FALSE, message = FALSE, warning = FALSE, fig.width=6, fig.height=8/1.6, out.width='600px', fig.align = 'center', tidy = TRUE, comment = NA, cache = FALSE, cache.path='cache/enrollment-', results = 'markup')
```

## Load required R packages and functions

```{r  loadpackages, results = 'hide', echo = TRUE}
options(digits = 3)
library(knitr)
library(ADNIMERGE)
library(Hmisc)
library(msSurv)
library(zoo)
library(ggplot2)
theme_set(theme_bw())
source("http://adni.bitbucket.io/myfunctions.R")

cbbPalette <- c('#0072B2', '#D55E00', '#CC79A7', '#E69F00', '#56B4E9', '#009E73', '#F0E442', '#000000')
scale_colour_discrete <- function(...) scale_colour_manual(..., values = cbbPalette) 
scale_fill_discrete <- function(...) scale_fill_manual(..., values = cbbPalette)
```

```{r  data, echo= FALSE}
mmse_bl <- mmse
mmse_bl$VISCODE <- with(mmse_bl, ifelse(VISCODE == "sc", "bl", VISCODE))
neurobat_bl <- subset(neurobat, VISCODE != "bl")
neurobat_bl$VISCODE <- with(neurobat_bl, ifelse(VISCODE == "sc", "bl", VISCODE))
cdr_bl <- cdr
cdr_bl$VISCODE <- with(cdr_bl, ifelse(VISCODE == "sc", "bl", VISCODE))
dd <- merge(dxsum[, c("RID", "COLPROT", "VISCODE", "DXCHANGE", 
  paste("DXMPTR", 1:6, sep = ""))], 
  mmse_bl[, c("RID", "COLPROT", "VISCODE", "MMSCORE")], 
  by = c("RID", "COLPROT", "VISCODE"), all = TRUE)
dd <- merge(dd, neurobat_bl[, c("RID", "COLPROT", "VISCODE", "LDELTOTAL")], 
  by = c("RID", "COLPROT", "VISCODE"), all = TRUE)
dd <- merge(dd, cdr_bl[, c("RID", "COLPROT", "VISCODE", "CDGLOBAL", "CDMEMORY")], 
  by = c("RID", "COLPROT", "VISCODE"), all = TRUE)
dd <- merge(dd, ecogpt[, c("RID", "COLPROT", "VISCODE", "CONCERN")], 
  by = c("RID", "COLPROT", "VISCODE"), all = TRUE)
dd <- merge(dd, registry[, c("RID", "COLPROT", "VISCODE", "EXAMDATE")], 
  by = c("RID", "COLPROT", "VISCODE"), all = TRUE)

dd <- merge(dd, subset(registry, VISCODE == "bl" & COLPROT == ORIGPROT, 
  c("RID", "COLPROT", "VISCODE", "EXAMDATE")), 
  by = "RID", all = TRUE, suffixes = c("", ".bl"))
dd <- merge(dd, subset(arm, COLPROT == ORIGPROT, c("RID", "DX")), 
  by = "RID", all = TRUE)
dd <- merge(dd, subset(ptdemog, VISCODE == "sc" & COLPROT == ORIGPROT, c("RID", "PTEDUCAT")), 
 by = "RID", all = TRUE)

COHORT <- unique(subset(registry, 
  (RGCONDCT == 'Yes' | VISTYPE == "Standard") & VISCODE == 'bl')$RID)
dd <- subset(dd, RID %in% COHORT & !VISCODE %in% c("sc", "scmri") & !is.na(DXCHANGE))

dd$Year <- as.numeric(with(dd, EXAMDATE - EXAMDATE.bl)/365.25)

dd$Dx <- as.character(dd$DXCHANGE)
dd$Dx <- gsub("Stable: ", "", dd$Dx)
dd$Dx <- gsub("Conversion: ", "", dd$Dx)
dd$Dx <- gsub("Reversion: ", "", dd$Dx)
dd$Dx <- gsub("MCI to ", "", dd$Dx)
dd$Dx <- gsub("Dementia to ", "", dd$Dx)
dd$Dx <- gsub("NL to ", "", dd$Dx)

dd$Dx.sc <- gsub("AD", "Dementia", dd$DX)

dd$obj_memory_loss <- apply(dd[, c('PTEDUCAT', 'LDELTOTAL')], 1, obj_memory)
dd$subj_memory_comp <- with(dd, 
  DXMPTR1 == "Yes" | DXMPTR2 == "Yes" | CONCERN == "Yes" | Dx.sc == "EMCI")

dd <- subset(dd, !is.na(Year))
dd <- dd[order(dd$RID, dd$Year), ]
# LOCF for ojective memory complaint since Log Memory II not done at m06 and m18
for(rid in unique(dd$RID[!is.na(dd$obj_memory_loss)])){
	dd[dd$RID == rid, "obj_memory_loss"] <- 
    na.locf(dd[dd$RID == rid, "obj_memory_loss"], na.rm = TRUE)
}

# MMSCORE %in% 24:30 & subj_memory_comp & 
# CDGLOBAL == 0.5 & CDMEMORY >= 0.5

dd$Dx2 <- with(dd, 
  ifelse(Dx == "MCI" & 
  obj_memory_loss %in% c("EMCI", "NL"), "EMCI", Dx))
dd$Dx2 <- ifelse(is.na(dd$Dx2), dd$Dx, dd$Dx2)
dd$stage4 <- as.numeric(as.character(factor(dd$Dx2, 
  levels = c("NL", "EMCI", "MCI", "Dementia"),
  labels = 1:4)))
# Dx3, stage3: no EMCI allowed at followup
dd$Dx3 <- with(dd, ifelse(Year == 0, Dx2, Dx))
dd$stage3 <- as.numeric(as.character(factor(dd$Dx3, 
  levels = c("NL", "EMCI", "MCI", "Dementia"),
  labels = 1:4)))

# remove duplicated RID EXAMDATEs
dd <- subset(dd, !duplicated(paste(RID, EXAMDATE)))

# first to final transition (including transitions to EMCI)
dd_first_to_final_4 <- with(dd, make_first_to_final_trans(RID, stage4, Year))
dd_all_int_final_4 <- with(dd, make_all_intermediate_trans(RID, stage4, Year))

# first to final transition (excluding transitions to EMCI)
dd_first_to_final_3 <- with(dd, make_first_to_final_trans(RID, stage3, Year))
dd_all_int_final_3 <- with(dd, make_all_intermediate_trans(RID, stage3, Year))

# Nodes <- c("1", "2", "3", "4")
# Edges <- list("1" = list(edges = c("2", "3", "4")),
#               "2" = list(edges = c("1", "3", "4")),
#               "3" = list(edges = c("1", "2", "4")),
#               "4" = list(edges = c("3")))
# treeobj <- new("graphNEL", nodes = Nodes, edgeL = Edges,
#             edgemode = "directed")
# fit4 <- msSurv(dd_all_int_final_4, treeobj, bs = TRUE, LT = TRUE)

Nodes <- c("1", "2", "3", "4")
Edges <- list("1" = list(edges = c("3", "4")),
              "2" = list(edges = c("1", "3", "4")),
              "3" = list(edges = c("1", "4")),
              "4" = list(edges = c("3")))
treeobj <- new("graphNEL", nodes = Nodes, edgeL = Edges,
            edgemode = "directed")
fit3 <- msSurv(dd_all_int_final_3, treeobj, bs = TRUE, LT = TRUE)
```

## Transition counts (all intermediate transitions)
```{r, echo= FALSE}
dd3 <- dd_all_int_final_3
dd3$end.stage <- with(dd3, ifelse(end.stage == 0, start.stage, end.stage))
dd3$start.stage <- factor(dd3$start.stage, levels = c(1:4), labels = c("NL", "EMCI", "LMCI", "AD"))
dd3$end.stage <- factor(dd3$end.stage, levels = c(1:4), labels = c("NL", "EMCI", "LMCI", "AD"))
addmargins(with(subset(dd3, end.stage != 0), table(start.stage, end.stage)))
```

## Transition counts (first to last stage transitions)
```{r}
ddf <- dd_first_to_final_3
ddf$start.stage <- factor(ddf$start.stage, levels = c(1:4), labels = c("NL", "EMCI", "LMCI", "AD"))
ddf$end.stage <- factor(ddf$end.stage, levels = c(1:4), labels = c("NL", "EMCI", "LMCI", "AD"))
addmargins(with(ddf, table(start.stage, end.stage)))
```

## Transition probabilities (%)
```{r  results = 'hide'}
dx <- c("NL", "EMCI", "LMCI", "AD")
tab1 <- Pst(fit3, s = 0, t = 1.1)
tab2 <- Pst(fit3, s = 0, t = 2.1)
tab3 <- Pst(fit3, s = 0, t = 3.1)
tab4 <- Pst(fit3, s = 0, t = 4.1)
```

### Year 0 to 1
```{r}
dimnames(tab1$Pst) <- list(dx, dx)
addmargins(tab1$Pst * 100)
```

### Year 0 to 2
```{r}
dimnames(tab2$Pst) <- list(dx, dx)
addmargins(tab2$Pst * 100)
```

### Year 0 to 3
```{r}
dimnames(tab3$Pst) <- list(dx, dx)
addmargins(tab3$Pst * 100)
```

### Year 0 to 4
```{r}
dimnames(tab4$Pst) <- list(dx, dx)
addmargins(tab4$Pst * 100)
```

## Transition probability plots
```{r}
my_msSurv_plot(fit3, plot.type="transprob", trans=c("1 2", "1 3", "1 4")  , 
  state.names = c("NL", "EMCI", "LMCI", "AD"),
  timelab = "Year")
my_msSurv_plot(fit3, plot.type="transprob", trans=c("2 1", "2 3", "2 4")  , 
  state.names = c("NL", "EMCI", "LMCI", "AD"),
  timelab = "Year")
my_msSurv_plot(fit3, plot.type="transprob", trans=c("3 1", "3 2", "3 4")  , 
  state.names = c("NL", "EMCI", "LMCI", "AD"),
  timelab = "Year")
```

## State occupation probabilities
```{r}
my_msSurv_plot(fit3, 
  state.names = c("NL", "EMCI", "LMCI", "AD"),
  timelab = "Year")
```

